---
title: "Reproducible Research: Peer Assessment 1"
author: "Daniel A Stearns"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
#### 1. Load Libraries
```{r, echo=TRUE}

library(data.table)
library(chron)

```

#### 2. Load the data 
```{r, echo=TRUE}

if(!file.exists('activity.csv')){
    unzip('activity.zip')
}

src_data <- read.csv("activity.csv")
```
#### 3. Review and the data set
```{r, echo=TRUE}

names(src_data)
str(src_data)
summary(src_data)
head(src_data)
tail(src_data)

```
#### 4. Preprocess the data
```{r, echo=TRUE}
src_data$date <- as.Date(as.character(src_data$date),format="%Y-%m-%d")

data <- data.table(src_data)
data <- data[, weekday := weekdays(as.Date(date))]
data <- data[, daytype := ifelse(is.weekend(data$date), "Weekend", "Weekday") ]

data_day <- data[, list(steps = sum(steps, na.rm = T)), by = date]
data_int <- data[, list(steps = mean(steps, na.rm = T)), by = interval]
data_cleansing_strategy <- data[, list(mean_steps = mean(steps, na.rm = T)), by =  list(interval,weekday)]
tables()

```

## What is mean total number of steps taken per day?
#### 1. Display Total Steps per Data
```{r, echo=TRUE}
data_day
```
#### 2. Plot Historgram of total steps per day
```{r, echo=TRUE}
hist(
   data_day$steps, 
   breaks=15, 
   main="Number of Days with Total Steps", 
   xlab="Daily Steps", 
   ylab="Frequency", 
   col="cyan"
)

#barplot(
#    data_day$steps,  
#    names=data_day$date, 
#    main="Total Steps Per Day", 
#    xlab="Dates", 
#    ylab="Steps", 
#    col="cyan", 
#    cex.names=0.6
#)

```

#### 3. Calculate Mean and Median

```{r, echo=TRUE}

median_daily_steps <- median(data_day[,steps])
mean_daily_steps <- mean(data_day[,steps])

```

#####Mean: `r format(median_daily_steps, scientific = FALSE)`
#####Median:  `r format(mean_daily_steps, scientific = FALSE)`

## What is the average daily activity pattern?
#### 1. Time series plot of the interval and steps taken.

```{r, echo=TRUE}

plot(
    data_int$interval, 
    data_int$steps, 
    type='l', 
    col="red", 
    main="Average Steps by Time of Day (5 Minute Intervals)", 
    xlab="Time of Day", 
    ylab="Steps"
)

```


#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo=TRUE}

max_step_interval <- data_int[which.max(data_int$steps)]$interval

```

#####Interval with Maximum Steps: `r format(max_step_interval, scientific = FALSE)`


## Inputing missing values
#### 1. Calculate Missing Values

```{r, echo=TRUE}

total_missing <- sum(is.na(data$steps))

```

#####Total Missing Values: `r total_missing`

#### 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Reviewing the data set we find we are missing 13.1% of the step values from the dataset.We would like to devise strategy for imputing the missing values, since the mean and median strategies are not the best options, I choose to create a strategy tha generates the mean value for steps based on the weekday and interval. I felt the weekday was an important factor as people have diffrent habits based on days of the week. If more data was present I may have also looked at seasonality of the steps. For example Fall, Winter, Spring, Summer. The imputed value is then applied to the NA values in the dataset where matched on the weekday and interval.

```{r, echo=TRUE}

data_clean <- data
data_clean <- merge(data_clean, data_cleansing_strategy, by=c("interval","weekday")) 
data_clean <- within(data_clean, steps <- ifelse(is.na(steps), mean_steps, steps))
data_clean <- data_clean[ ,list(date, interval, steps, daytype)]                     
                    

#verify there are no missing values
sum(is.na(data_clean$steps))

```

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r, echo=TRUE}
data_clean_day <- data_clean[, list(steps = sum(steps, na.rm = T)), by = date]

```

#### 4. Plot Historgram of total steps per day
```{r, echo=TRUE}
hist(
   data_clean_day$steps, 
   breaks=15, 
   main="Number of Days with Total Steps", 
   xlab="Daily Steps", 
   ylab="Frequency", 
   col="cyan"
)

#barplot(
#    data_clean_day$steps,  
#    names=data_clean_day$date, 
#    main="Total Steps Per Day", 
#    xlab="Dates", 
#    ylab="Steps", 
#    col="cyan", 
#    cex.names=0.6
#)

```

#### 3. Calculate Mean and Median

```{r, echo=TRUE}

median_clean_daily_steps <- median(data_clean_day[,steps])
mean_clean_daily_steps <- mean(data_clean_day[,steps])

```

#####Mean: `r format(median_clean_daily_steps, scientific = FALSE)`
#####Median:  `r format(mean_clean_daily_steps, scientific = FALSE)`

####4. Determine impact of imputing missing data.

#####Median Original: `r format(median_daily_steps, scientific = FALSE)` 
#####Median Clean: `r format(median_clean_daily_steps, scientific = FALSE)` 
#####<span style="color:blue">Median Difference: `r format(median_clean_daily_steps-median_daily_steps, scientific = FALSE)`</span>
#####<span style="color:blue">Median Variance: `r format(round((median_clean_daily_steps-median_daily_steps)/median_daily_steps*100,2), scientific = FALSE)`%</span>

#####Mean Original: `r format(mean_daily_steps, scientific = FALSE)` 
#####Mean Clean: `r format(mean_clean_daily_steps, scientific = FALSE)` 
#####<span style="color:blue">Mean Difference: `r format(mean_clean_daily_steps-mean_daily_steps, scientific = FALSE, digits=2)`</span> 
#####<span style="color:blue">Median Variance: `r format( round( (mean_clean_daily_steps - mean_daily_steps)/mean_daily_steps*100 ,2), scientific = FALSE)`%</span>

Yes, using the cleansing strategy based on average number of steps per interval and weekday causes a statistically significant shift in the median and mean amount of steps, however the imputing of the data creates a more normalized histogram with an even distribution. 


## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable 

```{r, echo=TRUE}

date_week <- data_clean
data_week <- data[, list(steps = mean(steps, na.rm = T)), by = list(interval, daytype)]
data_weekday <- data_week[daytype=="Weekend"]
data_weekend <- data_week[daytype=="Weekday"]

```

#### 2. Make a panel plot 

```{r, echo=TRUE}
par(mfrow=c(2,1))

plot(
    data_weekend$interval, 
    data_weekend$steps, 
    type='l', 
    col="red",
    main="Average Steps by Time of Day (5 Minute Intervals)

    Weekend",
    xlab="Time of Day", 
    ylab="Steps",
    ylim=c(0, 250)
)

plot(
    data_weekday$interval, 
    data_weekday$steps, 
    type='l', 
    col="red",
    main="Weekday", 
    xlab="Time of Day", 
    ylab="Steps",
    ylim=c(0, 250)
)


```

#### 3. Weekend/Weekday Differences 

```{r, echo=TRUE}
summary(data_weekday$steps)

summary(data_weekend$steps)
```